
---
title: "Spatio‑Temporal Patterns of African Hominins"
author: "Hominid Dates Project"
date: "`r format(Sys.Date(), '%d %b %Y')`"
format:
  pdf:
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
import::from("targets", tar_read)
import::from("dplyr", mutate, arrange)
import::from("kableExtra", kable_styling, kable_classic)
```

## Overview

This report provides a brief summary of specimen counts and temporal relationships in the **Hominid Dates** database.  
Full reproducible workflow available in the project GitHub repository (https://github.com/BWBrook/hominid-dates).  
Here we focus on headline patterns and intuitive graphics.  

---

## 1 Specimen totals

### 1.1 Genus counts

```{r genus-tbl}
genus_tbl <- tar_read(genus_species_freq)$genus_freq
knitr::kable(genus_tbl,
             caption = "Specimen counts per genus (collapsed by duplicate fragments).") |>
  kable_styling(full_width = FALSE) |>
  kable_classic()
```

```{r genus-plot, fig.cap="Specimen counts per genus"}
tar_read(genus_plot)
```

### 1.2 Species counts (top 15)

```{r species-tbl}
species_tbl <- tar_read(genus_species_freq)$species_freq
knitr::kable(species_tbl,
             caption = "Specimen counts per species (all taxa).") |>
  kable_styling(full_width = FALSE) |>
  kable_classic()
```

```{r species-plot, fig.cap="Top 15 species by specimen count"}
tar_read(species_plot)
```

---

## 2 Temporal ranges

`first_occ` marks the **oldest** directly or indirectly dated horizon for a species;  
`last_occ` marks the youngest. The line segments in Figure 3 visualise these bounds; points mark ends.

```{r spans-table}
span_tbl <- tar_read(temporal_span) |>
  arrange(desc(first_occ))
```

```{r spans-kable}
knitr::kable(span_tbl,
             caption = "First and last appearances by species (Ma).") |>
  kable_styling(latex_options = c("scale_down")) |>
  kable_classic()
```

```{r span-plot, fig.cap="Temporal ranges of each species"}
tar_read(span_plot)
```

---

## 3 Pairwise temporal overlap

Values express the length of time (in Ma) two species’ known intervals overlap.  
*Interpretation tips*:

* An overlap of **≈ 0 Ma** suggests sequential replacement or sampling gaps.  
* Larger cells (yellow in the heat‑map) denote prolonged co‑occurrence.  
* The diagonal is set to **NA** so that off‑diagonal relationships stand out; a species trivially overlaps itself by its full span.

```{r overlap-heatmap, fig.cap="Pairwise temporal overlap (Ma)"}
tar_read(overlap_plot)
```

---

## 4 Sampling intensity caveat

The pipeline collapses fragments to uniques but records a `sampling_intensity` count.  
Where sampling is sparse (low counts or long hiatuses), absences may reflect fossilisation and recovery bias rather than true local extirpation.  
Upcoming occupancy models will incorporate this intensity as a detection covariate.

---

## 5 Next steps

1. Implement Solow’s **OLE** to estimate continental extinction probabilities.  
2. Fit hierarchical occupancy models at key sites (e.g., Olduvai Gorge) integrating `sampling_intensity`.  
3. Visualise spatio‑temporal colonisation fronts via animated maps.
